---
title: "GSE22544 Compression"
output: html_document
---
This R script uses a uniqued gene symbol vector to loop through a expression dataframe to subset and sort out gene symbols in a list. This list is then changed in a dataframe/matrix, meaned, and eventually contains indiv gene symbols on the rownames and sample names on colnames. Contains meaned expression data.
```{r code to set working directory, eval=FALSE}
setwd("D:/Code/RE/My R scripts")
```

## (1) Loading up the combined dataset and gene symbol vector
```{r, eval=FALSE}
bound_GSE22544 <- read.csv("bound_GSE22544.csv", header = T, sep = ",", stringsAsFactors = FALSE)
bound_GSE22544 <- bound_GSE22544[,-1] # removes 1st column of numbers
bound_GSE22544 <- bound_GSE22544[!(is.na(bound_GSE22544$Gene.Symbol) | bound_GSE22544$Gene.Symbol=="")
                                 
gene_symbol_GSE22544 <- read.csv("gene_symbol_GSE22544.csv", header = T, sep = ",", stringsAsFactors = FALSE)
gene_symbol_GSE22544 <- gene_symbol_GSE22544[,2] # chooses only the second column
gene_symbol_GSE22544 <- na.omit(gene_symbol_GSE22544) # removes rows that have NAs
# View(gene_symbol_GSE22544) # 16,072 entries

```
## (2) Using unique() function to remove duplicate genes
```{r, eval=FALSE}
uniq_genesymbol <- unique(gene_symbol_GSE22544) #16,972 entries before unique function
# View(uniq_genesymbol)# 9,995 entries after unique function
```
## (3) Making an Empty List
```{r, eval=FALSE}
mat_list <- list()
```
## (4) Using a 'for' loop to loop through bound_GSE22544 using a gene symbol vector
```{r, eval=FALSE}
for (i in uniq_genesymbol){
  df_subset <- subset(bound_GSE22544, Gene.Symbol == i, select = -c(1))
  mat_subset <- as.matrix(df_subset)
  mat_list[[i]] <- mat_subset
}
```
## (5) Writing a function to find the mean for each sublist
```{r, eval=FALSE}
mean_fun <- function(x){
  mean_value <- colMeans(x)
  return(mean_value)
}

```
## (6) Using lapply on the list of gene symbols and exprs data using the function i created (mean_fun)

```{r, eval=FALSE}
mean_list <- lapply(mat_list, mean_fun)
# View(mean_list)
```
## (7) Exprs data compressed into mean exprs data
```{r, eval=FALSE}
# new_GSE22544 <- matrix(unlist(mean_list), ncol = 19, byrow = TRUE) # rejected cos no row and column names; only mean exprs data
# new_GSE22544 <- data.frame(matrix(unlist(mean_list), ncol = 19, byrow = TRUE)) # rejected cos no column and row names retained and needs extra steps to take the colnames from bound_22544 df
new_df_GSE22544 <- data.frame(t(sapply(mean_list,c))) # has gene symbols as rownames and sample numbers as column names
```
## Writing files
```{r,eval=FALSE}
write.csv(new_df_GSE22544, file = "D:/Code/RE/My R scripts/new_df_GSE22544.csv", row.names = T)
```


