---
title: "BS9001 Research Experience: GSE22544 Series Extraction Procedure"
author: Justin Chia Cheng Ze
date: July 31, 2019
output:
  prettydoc::html_pretty:
    theme: architect
    highlight: github
    
---
<style>
div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
</style>
<div class = "dark blue">

*This R script demonstrates the extraction of genetic expression data from the GEO database using R. Any data extracted is then assigned to variables for further use.*

</div>
### Set the working directory
The working directory is always set at the start of any R script to ensure that RStudio can read the files in the working directory.
```{r, eval=FALSE}
setwd("D:/Code/RE/My R scripts")
```

### (1) Obtain GSE series data from GEO database

Firstly, the data is extracted from a GEO GSE file using the `GEOquery` package:

```{r code for extracting GSE22544 data, message=FALSE, eval=FALSE}
library(GEOquery) # always use the library() function first in order to use a certain package
  
GSE22544 <- getGEO(filename = 'GSE22544_series_matrix.txt.gz')
GSE22544
      
```
The code has been commented to prevent the code from running.

### (2) Assigning pheno data and feature data into respective vectors
Then I assigned pheno data and feature data of the GSE into separate variables as such:
```{r code that assigns pheno data and feature data into variables, eval=FALSE}
      
pheno_GSE22544 <- phenoData(GSE22544) # pheno data
feature_GSE22544 <- fData(GSE22544) # feature data
```

### (3) Obtaining expression data from feature data using the `exprs()` function
```{r exprs data is extracted from feature data, eval=FALSE}
      exprs_GSE22544 <- exprs(GSE22544) # exprs() function applied to GSE22544
      class(exprs_GSE22544) # shows as matrix
      exprs_GSE22544 <- as.data.frame(exprs_GSE22544) # change matrix to data.frame
      # View(exprs_GSE22544) # GSM on the columns, 'at' on the rows
```

### (4) Subsetting the gene Symbols from feature data
```{r code that subsets the gene symbols, eval=FALSE}
      GENE_Symbols <- (feature_GSE22544 [11])
      rownames(GENE_Symbols) <- NULL # removes probe IDs as rownames
      class(GENE_Symbols)
```

### (5) Combine GENE_Symbols df with expression df
```{r using the cbind function to combine data frames, eval=FALSE}
      bound_GSE22544 <- cbind(GENE_Symbols, exprs_GSE22544)
      rownames(bound_GSE22544) <- NULL
      
      View(bound_GSE22544)
      ncol(bound_GSE22544)
```

### (6) Creating a class information vector and factor
```{r code to obtain class vector and factor, eval=FALSE}
# using gsub to change the class labels
pheno_GSE22544@data$characteristics_ch1 <- gsub("tissue:", "",   pheno_GSE22544@data$characteristics_ch1)
pheno_GSE22544@data$characteristics_ch1 <- gsub("breast cancer", "IDC", pheno_GSE22544@data$characteristics_ch1)
pheno_GSE22544@data$characteristics_ch1 <- gsub("normal breast", "normal", pheno_GSE22544@data$characteristics_ch1)
      
      # making a factor for use later in bootstrap step
      class_factor <- as.factor(pheno_GSE22544@data$characteristics_ch1)
      levels(class_factor)
      levels(class_factor) [1]
      levels(class_factor) [2]
      
      class_factor == levels(class_factor) [1] #  [1]  TRUE  TRUE  TRUE FALSE  TRUE  TRUE  TRUE FALSE FALSE  TRUE  TRUE  TRUE FALSE  TRUE  TRUE  TRUE  TRUE  TRUE  TRUE  TRUE
      
      
      class_factor == levels(class_factor) [2] #  [1] FALSE FALSE FALSE  TRUE FALSE FALSE FALSE  TRUE  TRUE FALSE FALSE FALSE  TRUE FALSE FALSE FALSE FALSE FALSE FALSE FALSE
      
      
      # making a vector for insertion into exprs dataframe later in bootstrap step
      class_vector <- c(pheno_GSE22544@data$characteristics_ch1)
```
### (7) Creating a gene symbols vector      
```{r code for gene symbols vector, eval=FALSE}
 gene_symbol_GSE22544 <- (feature_GSE22544 [11])
      rownames(gene_symbol_GSE22544) <- NULL
      # View(gene_symbol_GSE22544)
      
      gene_symbol_GSE22544 <- as.vector(gene_symbol_GSE22544[[1]])
      is.vector(gene_symbol_GSE22544)
      
      # writing into a csv file
      # write.csv()
```

### Writing files
```{r code for writing files, eval=FALSE}
write.csv(bound_GSE22544, file = "D:/Code/RE/My R scripts/bound_GSE22544.csv") # writes bound df with gene symbol and expression data
write.csv(gene_symbol_GSE22544, file = "D:/Code/RE/My R scripts/gene_symbol_GSE22544.csv") # writes gene symbol vector
write.csv(class_factor, file = "D:/Code/RE/My R scripts/class_factor.csv") # writes class labels factor
write.csv(class_vector, file = "D:/Code/RE/My R scripts/class_vector.csv")
```